---
title: "Analysis of variables affecting a car's fuel efficiency"
author: Robert Merriman
date: November 22, 2015
output: pdf_document
---
# **Executive Summary**
```{r, echo = F, results = "hide", message = F}
library(car)
library(MASS)
library(ggplot2)
library(sm)
library(gridExtra)
```
I work for Motor Trend, a magazine about the automobile industry. I analyzed the mtcars car data to answer the following two questions:

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions?

For the data analyzed, a car with a manual tranmission is likely to have a higher fuel efficiency (MPG) than a car with an automatic transmission. Using a liner regression, the transmission type did have a statistically significant increase (2.9358) in MPG but is confounded by other variables. As the data did not include any cars that were identical except for the transmission type, further analysis may be needed.

# **Methodology**
## **Source Data**
The source data is mtcars in the R datasets library. The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).^Ref1,FigureSet1^

## **Data Manipulation**
I converted the variables vs and am as factors and I relabled the am factors to Autmoatic and Manual to improve understanding the visuals. I did not convert the variables gear, cyl, or carb as factors as the variables are not dichotomous or ordinal. For example increasing the number of gears has enabled better engineering optimization of a car's fuel efficiency^Ref2^.
```{r, echo = F, results = "hide"}
M1 <- mtcars
M1$vs <- as.factor(M1$vs)
M1$am <- factor(M1$am, labels = c("Automatic", "Manual"))
mpg <- M1$mpg
meanMPG <- mean(mpg)
sdMPG <- sd(mpg)
```

## **Data Exploration**
In exploring the data I concluded that it is reasonable to consider MPG approximately normally distributed.  the distribtuion for cars with an automatic tranmission appears normally distributed with a long tail to the right.  The distribution for cars with a manual transmission are almost bimodal and are shifted to the right compared to cars with an automatic transmission. I noted that there are outliers at the right tail due to some cars with manual transmission having high MPG.^FigureSet2^

## **t.test of MPG ~ AM**
```{r, echo = F, results = "hide"}
tt <- t.test(mpg ~ am, data = mtcars)
```
The mean mileage of cars with an automatic transmission is `r round(mean(mtcars$mpg[mtcars$am == "0"],3))` mpg and for cars with a manual transmission is `r round(mean(mtcars$mpg[mtcars$am == "1"],3))` mpg. Treating the MPG data as i.i.d. and normally distributed, by t.test comparing MPG to the transmission type, the 95% confidence interval of the difference in mean gas mileage is between `r abs(round(tt$conf.int[2],3))` and `r abs(round(tt$conf.int[1],3))` mpg. As the mean does not include 0 and the p.value `r round(tt$p.value,4)` is less than an alpha level of 0.05, the difference is statistically significant.

# **Model Selection for linear regerssion**
I used stepAIC for the model selection. The best regression model was (mpg ~ wt + qsec + am) which did find transmission type to be statistically significant with qsec and weight being statistically significant confounders (at an alpha of 0.05). I believe that additional modeling approaches should be employed.^Ref3^ 
```{r, echo = F, results = "hide"}
# model selection via stepAIC
lmAll <- lm(mpg ~ ., data = M1)
bestmodel <- stepAIC(lmAll, direction="both")
```

### **Interpreting the linear model**^FigureSet3^

The plot of residuals vs fitted indicate that residuals are uncorrelated and therefore are homoscedastic with normally distributed errors. However the scale-location plot indicates that the residuals might be heteroscedastic which would require tuning of the model such as by centering & scaling variables, testing for the interaction between variables, and possibly using different variables. None of the residuals had high leverage. The Q-Q plot shows that the data is approximately normally distributed except at the right tail as previouly shown in the data exploration.^FigureSet2^


# **References**
1. [mtcars dataset](http://www.inside-r.org/r-doc/datasets/mtcars)
2. [Gear article](http://www.autoblog.com/2010/08/18/greenlings-why-do-automatic-transmissions-now-get-better-fuel-e/)
3. [fivethirtyeight](http://fivethirtyeight.com/features/science-isnt-broken/#part2)
4. [QQNorm](https://en.wikipedia.org/wiki/Q–Q_plot)
5. [Shapiro-Wilk normality test](https://en.wikipedia.org/wiki/Shapiro–Wilk_test)
6. [r-tutur.com](http://www.r-tutor.com/elementary-statistics/inference-about-two-populations/population-mean-between-two-independent-samples)
7. [Variances explained](http://varianceexplained.org/RData/code/code_lesson3/)
8. [Histogram Blog](http://asbidyarthy.blogspot.com/2012/01/histograms-and-density-plots.html)
9. Articles on interpreting lm: [LM1](http://stats.stackexchange.com/questions/52089/what-does-having-constant-variance-in-a-linear-regression-model-mean/52107#52107) [LM2](http://stats.stackexchange.com/questions/58141/interpreting-plot-lm/65864) [LM3](http://www.personality-project.org/r/r.lm.html) [LM4](http://blog.yhathq.com/posts/r-lm-summary.html)

# **Appendix**
## **Figure set 1: basic data exploration**
```{r, echo = F, warning = F, message = F}
summary(M1)
str(M1)
par(mfrow=c(1,1))
p1 <- ggplot(M1, aes(x=cyl, y=mpg, color=am)) + geom_point() + ggtitle("MPG by cyclinders") + xlab("Cylinders") + ylab("MPG")
p2 <- ggplot(M1, aes(x=wt, y=mpg, color=am)) + geom_point() + ggtitle("MPG by weight") + xlab("Weight") + ylab("MPG")
p3 <- ggplot(M1, aes(x=qsec, y=mpg, color=am)) + geom_point() + ggtitle("MPG by qsec") + xlab("Time (s) for quarter mile") + ylab("MPG")
p4 <- ggplot(data = M1, aes(x=factor(am), y=mpg)) + geom_boxplot() + ggtitle("MPG by transmission type") + xlab("Transmission type") + ylab("MPG")
grid.arrange(p1, p2, p3, p4)
```

## **Figure set 2: MPG distribution**
```{r, echo = F}
par(mfrow=c(1,2))

# frequency histogram
# histMPG<-hist(mpg, breaks = 10, density = 20, col = "lightgray", xlab = "MPG", main = "MPG with normal\ndistribution overlay") 
# xfit<-seq(min(mpg), max(mpg), length=40) 
# yfit<-dnorm(xfit, mean = meanMPG, sd = sdMPG) 
# yfit <- yfit*diff(histMPG$mids[1:2]) * length(mpg) 
# lines(xfit, yfit, col = "black", lwd = 2)

# density histogram
hist(mpg, density=20, breaks=10, prob=TRUE, 
     xlab = "MPG", ylim=c(0, .12), 
     main = "MPG probabilty histogram with\nnormal ndistribution overlay")
curve(dnorm(x, mean = meanMPG, sd = sdMPG), 
       col = "darkblue", lwd = 2, add = TRUE, yaxt = "n")

# # Kernel Density Plot
# plot(density(mtcars$mpg), xlab = "MPG", main = "MPG kernel density", ylim=c(0., .07))

# densities by am
sm.density.compare(M1$mpg, M1$am, xlab="mpg")
title(main="MPG kernel distribution\nby transmission")
colfill<-c(2:(2+length(levels(M1$am)))) 
legend("topright",levels(M1$am), fill=colfill) 
```

## **Figure set 3: stepAIC**
```{r, echo = F}
#residuals
summary(bestmodel)
par(mfrow=c(1,2))
plot(bestmodel)
```